Seabird assemblages are linked to the major western boundary current off eastern Australia

### 2023-10-17 ms_preprint/ms_preprint.Rmd
added specific formatting requirements for submission, such as 'title page' and all statements

---
title: Seabird assemblages are linked to the major western boundary current off eastern Australia
authors:
  - name: Nicholas W. Daudt
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: nicholaswdaudt@gmail.com
  - name: Eric J. Woehler
    department: Australasian Seabird Group
    affiliation: BirdLife Australia
    location: Hobart, Australia
  - name: Matthew R. Schofield
    department: Department of Mathematics and Statistics
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
  - name: Robert O. Smith
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
  - name: Leandro Bugoni
    department: Institute of Biological Sciences
    affiliation: Universidade Federal do Rio Grande--FURG
    location: Rio Grande, RS, Brazil
  - name: William J. Rayment
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
abstract: |
  **Aim**: To identify assemblages of seabird species and relate them to environmental conditions off eastern Australia, a world seabird hotspot.
  **Methods**: We used 15 seabird at-sea ship-based survey data collected over 37° of latitude, from 2016 to 2021, and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and abundance). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within regions, but different amongst them. These groups also vary according to covariates, which in our case included oceanographic, climatic, and physiographic parameters.
  **Results**: Based on 142,646 seabirds recorded, from 80 species, all models suggested two macro-scale assemblages (‘northern’ and ‘southern’), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at ~34°S, near the latitude at which the East Australian Current separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.
  **Conclusion**: RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given the rapid climatic changes that the East Australian Current is experiencing and its importance in influencing distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.
keywords:
  - Assemblages
  - Biogeography
  - Community ecology
  - East Australian Current
  - Mixture models
  - Region of Common Profiles
bibliography: ../references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
# output: rticles::arxiv_article
output:
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage{pdflscape}
- \DeclareUnicodeCharacter{0394}{\ensuremath{\Delta}} # needed to understand unicode for greek Delta in Table 1
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
---

@@

@@

The EAC originates from the South Equatorial Current (SEC), a tropical nutrient-poor current. When the SEC's North Caledonian Jet meets the Queensland Plateau (\~18°S), it splits into a north-flowing branch towards the Gulf of Papua and a south-flowing branch, which marks the start of the EAC [@kessler2013; @ganachaud2014]. The South Caledonian Jet completes the EAC system at \~22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (**Fig. \@ref(fig:fig1-study-area)a**; @kessler2013; @ganachaud2014). **SST/SSS characteristics? ROB?** The EAC interacts with the bathymetry of the seafloor, giving rise to oceanographic features and processes, such as smaller currents, upwellings, and persistent dynamic eddies [@oke2000; @ridgway2003a; @oke2019]. The EAC flow polewards and is primarily associated with the shelf break, but at \~32°S it detaches from the shelf and spawns an area of eddy generation that dissipates throughout the western Tasman Sea [@mata2006; @oke2019], the 'EAC eastern extension' (@oke2019a; often termed the 'Tasman Front'). Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the 'EAC southern extension'; **Fig. \@ref(fig:fig1-study-area)a**), carrying eddy-like features as a transient current [@oke2019].

@@

Seabird data were gathered aboard *RV Investigator* and *RV Falkor* on 15 voyages, between 2016/17--2020/21, spanning over \~37° of latitude (**Fig. \@ref(fig:fig1-study-area)b**; **Table S1**). One or more observers continuously recorded seabirds within a 300 m forward quadrant (eye height 25 m a.s.l) from sunrise to sunset hours while the ship was underway. The species, the numbers of birds, and their behaviours were logged in real time into a ship-based web server. Environmental parameters for atmospheric and oceanographic conditions were logged continuously by automated-ship based sensors, with seabird observations supplementing relevant environmental conditions such as visibility or associations with marine mammals. Individual records received a date/time/geographic coordinate label (multi-species records have the same label but are logged as separate entries for each species). Birds following the vessels were excluded from analyses. Seabird observation protocols followed @raymond2010, in use around Australia since 1980.

@@

Seabirds are highly mobile animals, and many species migrate and/or disperse during their non-breeding periods [@biology2002]. Therefore, seasonal differences in species composition were expected [e.g. @reid2023], and in the probabilities of occurrence and in total numbers. Thus, we built separate modelling dataset for each season. We aggregated seabird data by 1° latitude $\times$ 1° longitude grid cells (c. 110 $\times$ 110 km) and season (**Fig. \@ref(fig:fig1-study-area)b**), by summing seabirds from the same species. To test if the presence-absence and abundance (raw counts) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \>1 to 1. To avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons, and environmental data were standardised before modelling.

@@

A total of 10,261 seabird records was obtained during the 15 voyages, comprising 142,646 individuals from 80 species, belonging to 11 families and five orders. The five most abundant species made up 83% of total birds counted (43.0% short-tailed shearwater *Ardenna tenuirostris*, 17.2% fairy prion *Pachyptila turtur*, 14.3% wedge-tailed shearwater *Ardenna pacifica*, 5.2% sooty tern *Onychoprion fuscatus*, 3.2% shy albatross *Thalassarche cauta*). The five most frequently encountered species were *A. pacifica* (frequency of occurrence = 35.3%), *T. cauta* (33.2%), brown booby *Sula leucogaster* (25.4%), red-footed booby *S. sula* (25.4%) and black-browed albatross *T. melanophris* (25%) (**Fig. S1**). An increase in species richness towards the south of the study region was clear in all but the winter season (**Fig. S2**). Noticeably higher counts were made around Tasmania, possibly due to the 'colony effect' and/or the presence of Southern Ocean species (**Fig. S2**). However, one cell in the Coral Sea had counts of the same order of magnitude as off Tasmania in summer.

@@

For all models, two RCPs was the optimal number of seabird groups, except for the presence-absence-autumn model, which indicated three RCPs as the best fit (**Fig. S3**). The best fitting models generally had 1 or 2 covariates selected, although the abundance-autumn model had five covariates (**Table \@ref(tab:table2)**). The best models showed a good fit with no apparent deviations from the assumptions (**Fig. S4**). All the best models selected 'sst' or 'sss', as essential for characterising water masses. The probability of belonging to one or another RCP varied its 'sst' threshold between 18--21 °C, or 'sss' 35.1 PSU for summer (**Fig. \@ref(fig:fig2-partial-plots)**, **Fig. S5**), similar to the EAC's temperature and salinity thresholds. Furthermore, most of the best models selected 'bat', suggesting a gradient between 'coastal' and 'oceanic' taxa based on bathymetry (**Fig. S5**). Of particular note, proxies for persistent oceanographic features that may aggregate food resources (the 'climatic' variables) for seabirds were selected only in the autumn and spring abundance models.

Grid point-predictions are the assignment of the RCP with the highest probability value for a given cell. When predicting RCP onto space, patterns of point-predictions suggest a biogeographic transition in seabird species around \~34°S, with both presence-absence and abundance models pointing to similar results (**Fig. \@ref(fig:fig3-point-pred)**). These patterns are further reinforced when looking at the mean probability along with the lower and upper confidence intervals of predictions (**Fig. \@ref(fig:fig4-prob-maps)**, **Fig. S6**), showing that around \~34°S, the 'northern' RCP tended to have higher probabilities in coastal regions, whereas the 'southern' RCP spread northward offshore. The presence-absence-autumn model (three RCPs) had a similar broad-scale pattern, although the southern RCP is farther south, and a strong 'Tasmania/Bass Straight' assemblage appears (**Fig. \@ref(fig:fig3-point-pred)**, **Fig. S6**).

The broad pattern of two large-scale assemblages is reflected in their species compositions. Although there are overlaps among the seabird groups, both presence-absence and abundance models captured typical tropical species in the northern RCP (e.g. boobies *Sula* spp., lesser frigatebird *Fregata ariel*) and typical Southern Ocean species in the southern RCP (e.g. small albatrosses *Thalassarche* spp., fairy prion *P. turtur*) (**Fig. \@ref(fig:fig5-spp-profiles)**, **Fig. S7**). The presence-absence-autumn model, which had three RCPs, also showed the characteristic taxa representing the 'northern' and 'southern' RCPs. However, the intermediate 'Tasmania/Bass Straight' assemblage showed a mixture of both groups with shy albatross and Australasian gannet *Morus serrator* prominent with high probabilities of occurrence.

@@

The seasonal models also captured the influence of migratory species, particularly in the southern assemblage. The summer and autumn models reflected the high abundance of short-tailed shearwater before they depart on their trans-equatorial migration [@carey2014]; the presence of the Australasian gannet was picked up by the autumn and winter models [@ismar2011]; and the white-headed petrel *Pterodroma lessonii* movements during chick-rearing and pre-laying exodus [@taylor2021] were detected in autumn and spring models, respectively. Given the geographic location of the southern assemblage, however, we believe sub-Antarctic taxa were not well represented in our models due to their low number of occurrences and relative paucity of survey effort to date (**Fig. S1**; hence they were removed before modelling). Additional surveys may reveal the seasonal northward pulse of these taxa into the southern assemblage (e.g. albatrosses *Diomedea* spp. and *Phoebetria* spp.), given that more effort translates as higher likelihood of recording these species. In turn, including these taxa in the models could result in a more stable 'three assemblage pattern' (as revealed by the autumn presence-absence model) in all but the summer season, when the majority of Antarctic species are breeding [@marchant1990handbook].

@@

# Acknowledgements

@@

@@

-   **TABLE 1.** Environmental data used in the Region of Common Profile models applied to seabird data off eastern Australia, their spatial/temporal resolutions, units, and sources
-   **TABLE 2.** Full (a) and best (b) Region of Common Profile Models applied to seabird data off eastern Australia, for both data types used (presence-absence, based on a Bernoulli model; abundance, based on a Negative Binomial model). Refer to Table 1 for the environmental data acronyms
-   **FIGURE 1.** Study area showing the main ocean currents (a), and seabird records made by season on top of the 1° latitude × 1° longitude grid cells (b). In (a), the East Australian Current (EAC) system is highlighted with its acronyms in bold and drawn in blue. The South Equatorial Current (SEC) and its branches, the North Caledonian Jet (NCJ) and South Caledonian Jet (SCJ), are drawn in red. The EAC 'eastern extension' is an eddy field (blue shaded area) often called the 'Tasman Front', from which the East Auckland Current originates (EAUC, in green). Figure (a) is adapted after Oke, Roughan, et al. (2019); for colour references see the online version
-   **FIGURE 2.** Partial plots for the retained covariates in the spring (a) presence-absence and (b) abundance best models. The plot shows the fitted probability of belonging to a Region of Common Profiles (RCP) against the environmental value. All other predictors are held at their mean values. Results for other seasons and models can be found in the Supplementary Material (Fig. S5). Refer to Table 1 for the environmental data acronyms
-   **FIGURE 3.** Spatialized point predictions (hard-classes) for seabird assemblages (Region of Common Profile; RCP), by season, based on presence-absence (a) and abundance (b) models, off eastern Australia. Each grid was assigned to the RCP group with highest membership probability
-   **FIGURE 4.** Predicted probability membership of each seabird assemblage (Region of Common Profiles; RCP) and grid, off eastern Australia, from spring presence-absence (a) and abundance (b) models. The central column, 'mean', corresponds to the point prediction and Bayesian boot-strapped, lower and upper confidence intervals (CI), on its sides. Note that, in case of two RCPs, RCP2 probablility equals to (1 - RCP1 probability). Results for other seasons and models can be found in the Supplementary Material (Fig. S6)
-   **FIGURE 5.** Species profiles of seabird assemblages (Region of Common Profile; RCP) off eastern Australia in spring. Values are the average and 95% confidence intervals of probability of occurrence (a) and predicted mean abundance (b) for each species, based on 1000 Bayesian bootstraps. Values in (b) were log10-transformed to accommodate the high variation between species. Results for other seasons and models can be found in the Supplementary Material (Fig. S7)

>>>>

---
title: Seabird assemblages are linked to the major western boundary current off eastern Australia
authors:
  - name: Nicholas W. Daudt*
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: nicholaswdaudt@gmail.com
    notes: Corresponding author
  - name: Eric J. Woehler
    department: Australasian Seabird Group
    affiliation: BirdLife Australia
    location: Hobart, Australia
    email: eric.woehler@gmail.com
  - name: Matthew R. Schofield
    department: Department of Mathematics and Statistics
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: matthew.schofield@otago.ac.nz
  - name: Robert O. Smith
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: robert.smith@otago.ac.nz
  - name: Leandro Bugoni
    department: Institute of Biological Sciences
    affiliation: Universidade Federal do Rio Grande--FURG
    location: Rio Grande, RS, Brazil
    email: lbugoni@yahoo.com.br
  - name: William J. Rayment
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: will.rayment@otago.ac.nz
abstract: |
  **Aim:** To identify assemblages of seabird species and relate them to environmental conditions off eastern Australia, a world seabird hotspot.
  **Location:** East coast of Australia.
  **Methods:** We used 15 seabird at-sea ship-based survey data collected over 37° of latitude, from 2016 to 2021, and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and abundance). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within regions, but different amongst them. These groups also vary according to covariates, which in our case included oceanographic, climatic, and physiographic parameters.
  **Results:** Based on 142,646 seabirds recorded, from 80 species, all models suggested two macro-scale assemblages (‘northern’ and ‘southern’), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at ~34°S, near the latitude at which the East Australian Current separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.
  **Main conclusions:** RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given the rapid climatic changes that the East Australian Current is experiencing and its importance in influencing distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.
keywords:
  - Assemblages
  - Biogeography
  - Community ecology
  - East Australian Current
  - Mixture models
  - Region of Common Profiles
bibliography: ../references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
output:
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage{pdflscape}
- \DeclareUnicodeCharacter{0394}{\ensuremath{\Delta}} # needed to understand unicode for greek Delta in Table 1
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
---

@@

# ORCID {.unnumbered}

Nicholas W. Daudt <https://orcid.org/0000-0001-8211-8771>

Eric J. Woehler <https://orcid.org/0000-0002-1125-0748>

Matthew R. Schofield <https://orcid.org/0000-0003-1481-2766>

Robert O. Smith <https://orcid.org/0000-0001-5931-3843>

Leandro Bugoni <https://orcid.org/0000-0003-0689-7026>

William J. Rayment -- **do you have ORCID, Will?**

# Biosketch {.unnumbered}

**Nicholas W. Daudt (he/him)** is broadly interested in spatial ecology and ecological modelling. His research uses data science and modelling to understand biogeographic patterns and ecological processes that build marine top predators' biodiversity patterns across space and time.

# Acknowledgements {.unnumbered}

***

**Data availability statement:** The gridded data used as the basis for all analyses is available in the archived repository along with the code to reproduce the results. The raw data are currently being used for other projects and thus can not be publicly available. If you are interested in the raw data, please get in touch with EJW.

**Funding statement:** .........**Eric - any specific funding(s) we need to acknowledge?**

**Conflict of interest disclosure:** The authors declare no conflict of interest.

**Ethics approval statement:** NA.

**Patient consent statement:** NA.

**Permission to reproduce material from other sources:** NA.

**Clinical trial registration:** NA.

\begin{center}
\textbf{Seabird assemblages are linked to the major western boundary current off eastern Australia}
\end{center}

*Running title:* Seabird assemblages off Eastern Australia

***

**Abstract**

**Aim:** To identify assemblages of seabird species and relate them to environmental conditions off eastern Australia, a world seabird hotspot.

**Location:** East coast of Australia.

**Methods:** We used 15 seabird at-sea ship-based survey data collected over 37° of latitude, from 2016 to 2021, and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and abundance). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within regions, but different amongst them. These groups also vary according to covariates, which in our case included oceanographic, climatic, and physiographic parameters.

**Results:** Based on 142,646 seabirds recorded, from 80 species, all models suggested two macro-scale assemblages (‘northern’ and ‘southern’), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at ~34°S, near the latitude at which the East Australian Current separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.

**Main conclusions:** RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given the rapid climatic changes that the East Australian Current is experiencing and its importance in influencing distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.

***Keywords*** Assemblages; Biogeography; Community ecology; East Australian Current; Mixture models; Region of Common Profiles

@@

The EAC originates from the South Equatorial Current (SEC), a tropical nutrient-poor current. When the SEC's North Caledonian Jet meets the Queensland Plateau (\~18°S), it splits into a north-flowing branch towards the Gulf of Papua and a south-flowing branch, which marks the start of the EAC [@kessler2013; @ganachaud2014]. The South Caledonian Jet completes the EAC system at \~22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (**Figure \@ref(fig:fig1-study-area)a**; @kessler2013; @ganachaud2014). The EAC interacts with the bathymetry of the seafloor, giving rise to oceanographic features and processes, such as smaller currents, upwellings, and persistent dynamic eddies [@oke2000; @ridgway2003a; @oke2019]. The EAC flow polewards and is primarily associated with the shelf break, but at \~32°S it detaches from the shelf and spawns an area of eddy generation that dissipates throughout the western Tasman Sea [@mata2006; @oke2019], the 'EAC eastern extension' (@oke2019a; often termed the 'Tasman Front'). Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the 'EAC southern extension'; **Figure \@ref(fig:fig1-study-area)a**), carrying eddy-like features as a transient current [@oke2019].

@@

Seabird data were gathered aboard *RV Investigator* and *RV Falkor* on 15 voyages, between 2016/17--2020/21, spanning over \~37° of latitude (**Figure \@ref(fig:fig1-study-area)b**; **Table S1**). One or more observers continuously recorded seabirds within a 300 m forward quadrant (eye height 25 m a.s.l) from sunrise to sunset hours while the ship was underway. The species, the numbers of birds, and their behaviours were logged in real time into a ship-based web server. Environmental parameters for atmospheric and oceanographic conditions were logged continuously by automated-ship based sensors, with seabird observations supplementing relevant environmental conditions such as visibility or associations with marine mammals. Individual records received a date/time/geographic coordinate label (multi-species records have the same label but are logged as separate entries for each species). Birds following the vessels were excluded from analyses. Seabird observation protocols followed @raymond2010, in use around Australia since 1980.

@@

Seabirds are highly mobile animals, and many species migrate and/or disperse during their non-breeding periods [@biology2002]. Therefore, seasonal differences in species composition were expected [e.g. @reid2023], and in the probabilities of occurrence and in total numbers. Thus, we built separate modelling dataset for each season. We aggregated seabird data by 1° latitude $\times$ 1° longitude grid cells (c. 110 $\times$ 110 km) and season (**Figure \@ref(fig:fig1-study-area)b**), by summing seabirds from the same species. To test if the presence-absence and abundance (raw counts) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \>1 to 1. To avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons, and environmental data were standardised before modelling.

@@

A total of 10,261 seabird records was obtained during the 15 voyages, comprising 142,646 individuals from 80 species, belonging to 11 families and five orders. The five most abundant species made up 83% of total birds counted (43.0% short-tailed shearwater *Ardenna tenuirostris*, 17.2% fairy prion *Pachyptila turtur*, 14.3% wedge-tailed shearwater *Ardenna pacifica*, 5.2% sooty tern *Onychoprion fuscatus*, 3.2% shy albatross *Thalassarche cauta*). The five most frequently encountered species were *A. pacifica* (frequency of occurrence = 35.3%), *T. cauta* (33.2%), brown booby *Sula leucogaster* (25.4%), red-footed booby *S. sula* (25.4%) and black-browed albatross *T. melanophris* (25%) (**Figure S1**). An increase in species richness towards the south of the study region was clear in all but the winter season (**Figure S2**). Noticeably higher counts were made around Tasmania, possibly due to the 'colony effect' and/or the presence of Southern Ocean species (**Figure S2**). However, one cell in the Coral Sea had counts of the same order of magnitude as off Tasmania in summer.

@@

For all models, two RCPs was the optimal number of seabird groups, except for the presence-absence-autumn model, which indicated three RCPs as the best fit (**Figure S3**). The best fitting models generally had 1 or 2 covariates selected, although the abundance-autumn model had five covariates (**Table \@ref(tab:table2)**). The best models showed a good fit with no apparent deviations from the assumptions (**Figure S4**). All the best models selected 'sst' or 'sss', as essential for characterising water masses. The probability of belonging to one or another RCP varied its 'sst' threshold between 18--21 °C, or 'sss' 35.1 PSU for summer (**Figure \@ref(fig:fig2-partial-plots)**, **Figure S5**), similar to the EAC's temperature and salinity thresholds. Furthermore, most of the best models selected 'bat', suggesting a gradient between 'coastal' and 'oceanic' taxa based on bathymetry (**Figure S5**). Of particular note, proxies for persistent oceanographic features that may aggregate food resources (the 'climatic' variables) for seabirds were selected only in the autumn and spring abundance models.

Grid point-predictions are the assignment of the RCP with the highest probability value for a given cell. When predicting RCP onto space, patterns of point-predictions suggest a biogeographic transition in seabird species around \~34°S, with both presence-absence and abundance models pointing to similar results (**Figure \@ref(fig:fig3-point-pred)**). These patterns are further reinforced when looking at the mean probability along with the lower and upper confidence intervals of predictions (**Figure \@ref(fig:fig4-prob-maps)**, **Figure S6**), showing that around \~34°S, the 'northern' RCP tended to have higher probabilities in coastal regions, whereas the 'southern' RCP spread northward offshore. The presence-absence-autumn model (three RCPs) had a similar broad-scale pattern, although the southern RCP is farther south, and a strong 'Tasmania/Bass Straight' assemblage appears (**Figure \@ref(fig:fig3-point-pred)**, **Figure S6**).

The broad pattern of two large-scale assemblages is reflected in their species compositions. Although there are overlaps among the seabird groups, both presence-absence and abundance models captured typical tropical species in the northern RCP (e.g. boobies *Sula* spp., lesser frigatebird *Fregata ariel*) and typical Southern Ocean species in the southern RCP (e.g. small albatrosses *Thalassarche* spp., fairy prion *P. turtur*) (**Figure \@ref(fig:fig5-spp-profiles)**, **Figure S7**). The presence-absence-autumn model, which had three RCPs, also showed the characteristic taxa representing the 'northern' and 'southern' RCPs. However, the intermediate 'Tasmania/Bass Straight' assemblage showed a mixture of both groups with shy albatross and Australasian gannet *Morus serrator* prominent with high probabilities of occurrence.

@@

The seasonal models also captured the influence of migratory species, particularly in the southern assemblage. The summer and autumn models reflected the high abundance of short-tailed shearwater before they depart on their trans-equatorial migration [@carey2014]; the presence of the Australasian gannet was picked up by the autumn and winter models [@ismar2011]; and the white-headed petrel *Pterodroma lessonii* movements during chick-rearing and pre-laying exodus [@taylor2021] were detected in autumn and spring models, respectively. Given the geographic location of the southern assemblage, however, we believe sub-Antarctic taxa were not well represented in our models due to their low number of occurrences and relative paucity of survey effort to date (**Figure S1**; hence they were removed before modelling). Additional surveys may reveal the seasonal northward pulse of these taxa into the southern assemblage (e.g. albatrosses *Diomedea* spp. and *Phoebetria* spp.), given that more effort translates as higher likelihood of recording these species. In turn, including these taxa in the models could result in a more stable 'three assemblage pattern' (as revealed by the autumn presence-absence model) in all but the summer season, when the majority of Antarctic species are breeding [@marchant1990handbook].

@@

@@

# Data Accessibility Statement {.unnumbered}

The gridded data used as the basis for all analyses is available in the archived repository along with the code to reproduce the results. The raw data are currently being used for other projects and thus can not be publicly available. If you are interested in the raw data, please get in touch with XXX.

@@

-   **TABLE 1.** Environmental data used in the Region of Common Profile models applied to seabird data off eastern Australia, their spatial/temporal resolutions, units, and sources
-   **TABLE 2.** Full (a) and best (b) Region of Common Profile Models applied to seabird data off eastern Australia, for both data types used (presence-absence, based on a Bernoulli model; abundance, based on a Negative Binomial model). Refer to Table 1 for the environmental data acronyms
-   **FIGURE 1.** Study area showing the main ocean currents (a), and seabird records made by season on top of the 1° latitude × 1° longitude grid cells (b). In (a), the East Australian Current (EAC) system is highlighted with its acronyms in bold and drawn in blue. The South Equatorial Current (SEC) and its branches, the North Caledonian Jet (NCJ) and South Caledonian Jet (SCJ), are drawn in red. The EAC 'eastern extension' is an eddy field (blue shaded area) often called the 'Tasman Front', from which the East Auckland Current originates (EAUC, in green). Figure (a) is adapted after Oke, Roughan, et al. (2019); for colour references see the online version
-   **FIGURE 2.** Partial plots for the retained covariates in the spring (a) presence-absence and (b) abundance best models. The plot shows the fitted probability of belonging to a Region of Common Profiles (RCP) against the environmental value. All other predictors are held at their mean values. Results for other seasons and models can be found in the Supplementary Material (Figure S5). Refer to Table 1 for the environmental data acronyms
-   **FIGURE 3.** Spatialized point predictions (hard-classes) for seabird assemblages (Region of Common Profile; RCP), by season, based on presence-absence (a) and abundance (b) models, off eastern Australia. Each grid was assigned to the RCP group with highest membership probability
-   **FIGURE 4.** Predicted probability membership of each seabird assemblage (Region of Common Profiles; RCP) and grid, off eastern Australia, from spring presence-absence (a) and abundance (b) models. The central column, 'mean', corresponds to the point prediction and Bayesian boot-strapped, lower and upper confidence intervals (CI), on its sides. Note that, in case of two RCPs, RCP2 probablility equals to (1 - RCP1 probability). Results for other seasons and models can be found in the Supplementary Material (Figure S6)
-   **FIGURE 5.** Species profiles of seabird assemblages (Region of Common Profile; RCP) off eastern Australia in spring. Values are the average and 95% confidence intervals of probability of occurrence (a) and predicted mean abundance (b) for each species, based on 1000 Bayesian bootstraps. Values in (b) were log10-transformed to accommodate the high variation between species. Results for other seasons and models can be found in the Supplementary Material (Figure S7)
